Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-05-06 paper/graph.tex
Minor changes to paper.  Some new results.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@608 3fc257bb-8550-0410-81b3-a5c103da83f7

Note that the degree distribution of vertices in the graph depends on
the choice of parameters $a$, $b$, $c$, $d$.  If one of the four
values is larger than the other three, a highly skewed distribution
results.  Variants of the above algorithm can be used when $N$ is not
a power-of-two, to generate graphs with weighted edges (assign a
numeric value to the $A_{ij}$ edge), graphs without self edges
(require $i \ne j$), or graphs with undirected edges (require $i < j$).
Or the general R-MAT matrix can be further processed by MapReduce
operations to meet these requirements.

By using more than one MapReduce object, we can improve the performance
of the R-MAT generation algorithm.  In the enhanced algorithm
(Figure~\ref{fig:rmat2}), we emit newly created edges into MapReduce object
$E_{new}$ and {\it aggregate()} the edges to processors.  
We then add those edges
to the MapReduce object $E_{old}$, containing all previously generated
unique edges; the edges of $E_{old}$ are already aggregated to processors
using the same mapping as $E_{new}$.  We can then use the {\it compress()}
function to reduce duplicates locally on each processor.
Performance improvements
in this algorithm result from the reduced number of KV pairs communicated in
{\it aggregate()},
compared to the collate operation in the original algorithm, 
particularly in later iterations of the generation where few
edges are created.

@@

Map: assign random values to each vertex \\
Clone: convert edge key/values directly to key/multivalues \\
Create: empty MapReduce object MRv for maximal independent set vertices \\
\\
while $N_{edges} > 0$: \\
\> 1 Reduce: \> Determine WINNER/LOSER vertex of each edge \\
       \> \> \> input Key = $E_{ij}$ with $V_i < V_j$, Multivalue = NULL or LOSER \\
       \> \> \> if either value is LOSER, emit nothing \\
       \> \> \> $V_w$ = WINNER vertex, $V_l$ = LOSER vertex \\
       \> \> \> output Key = $V_w$, Value = $V_l$ WINNER \\
       \> \> \> output Key = $V_l$, Value = $V_w$ LOSER \\
\> 1 Collate: \> Vertex as key \\
\> 2 Reduce: \> Find WINNER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are WINNERs: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 2 Collate: \> Vertex as key \\
\> 3 Reduce: \> Find LOSER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if any $V$ is LOSER: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 3 Collate: \> Vertex as key \\
\> 4 Reduce: \> Emit WINNER vertices and LOSER edges \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are LOSERs: \\
      \> \> \> \> output to MRv, Key = $V_i$, Value = NULL \\
       \> \> \> for each $V_j$ in Multivalue: \\
      \> \> \> \> if $V_j$ is LOSER: \\
     \> \> \> \> \> output Key = $E_{ij}$ with $V_i < V_j$, Value = LOSER \\
      \> \> \> \> else: \\
     \> \> \> \> \> output Key = $E_{ij}$ with $V_i < V_j$, Value = NULL \\
\> 4 Collate: \> Edge as key \\

>>>>

Note that the degree distribution of vertices in the graph depends on
the choice of parameters $a$, $b$, $c$, $d$.  If one of the four
values is larger than the other three, a highly skewed distribution
results.  Variants of the above algorithm can be used when $N$ is not
a power-of-two, to generate graphs with weighted edges (assign a
numeric value to the $A_{ij}$ edge), graphs without self edges
(require $i \ne j$), or graphs with undirected edges (require $i < j$).
Or the general R-MAT matrix can be further processed by MapReduce
operations to meet these requirements.  For some algorithms below
(e.g., triangle finding and maximal independent set generation), we
post-process R-MAT matrices to create upper-triangular R-MAT
matrices representing undirected graphs.  For each edge $(i,j)$ 
with $i>j$, we add edge $(j,i)$, essentially symmetrizing the matrix; 
we then remove duplicate edges and self edges,
leaving only edges with $i<j$.

By using more than one MapReduce object, we can improve the performance
of the R-MAT generation algorithm, particularly when out-of-core operations
are needed.  In the enhanced algorithm
(Figure~\ref{fig:rmat2}), we emit newly created edges into MapReduce object
$E_{new}$ and {\it aggregate()} the edges to processors.  
We then add those edges
to the MapReduce object $E_{old}$, containing all previously generated
unique edges; the edges of $E_{old}$ are already aggregated to processors
using the same mapping as $E_{new}$.  We can then use the {\it compress()}
function to reduce duplicates locally on each processor.
Time reductions
in this algorithm result from the reduced number of KV pairs communicated in
{\it aggregate()},
compared to the collate operation in the original algorithm, 
particularly in later iterations of the generation where few
edges are created.

@@

Map: assign random values to each vertex $V_i$ and $V_j$ of edge $E_{ij}$. \\
Clone: convert edge key/values directly to key/multivalues \\
Create: empty MapReduce object MRv for maximal independent set vertices \\
\\
while $N_{edges} > 0$: \\
\> 1 Reduce: \> Determine WINNER/LOSER vertex of each edge \\
       \> \> \> input Key = $E_{ij}$ with $V_i < V_j$, Multivalue = NULL or LOSER \\
       \> \> \> if either value is LOSER, emit nothing \\
       \> \> \> $V_w$ = WINNER vertex, $V_l$ = LOSER vertex \\
       \> \> \> output Key = $V_w$, Value = $V_l$ WINNER \\
       \> \> \> output Key = $V_l$, Value = $V_w$ LOSER \\
\> 1 Collate: \> Vertex as key \\
\> 2 Reduce: \> Find WINNER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are WINNERs: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 2 Collate: \> Vertex as key \\
\> 3 Reduce: \> Find LOSER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if any $V$ is LOSER: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V_j$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 3 Collate: \> Vertex as key \\
\> 4 Reduce: \> Emit WINNER vertices and LOSER edges \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are LOSERs: \\
      \> \> \> \> output to MRv, Key = $V_i$, Value = NULL \\
       \> \> \> for each $V_j$ in Multivalue: \\
      \> \> \> \> if $V_j$ is LOSER: \\
     \> \> \> \> \> output Key = $E_{ij}$ with $V_i < V_j$, Value = LOSER \\
      \> \> \> \> else: \\
     \> \> \> \> \> output Key = $E_{ij}$ with $V_i < V_j$, Value = NULL \\
\> 4 Collate: \> Edge as key \\
